ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-11-16 pyked.tex
removed formatting for final revision; this is the final version for IJCK

% For the version with changes, use these four commands
% *** Be very precise and careful about including whitespace and punctuation in your edits ***
\usepackage[normalem]{ulem}
\newcommand{\addone}[1]{{\sloppy\textcolor{teal}{\uline{#1}}}}  % add
\newcommand{\deleteone}[1]{\sloppy\textcolor{teal}{\sout{#1}}}    % delete
\newcommand{\addtwo}[1]{{\sloppy\textcolor{red}{\uline{#1}}}}  % add
\newcommand{\deletetwo}[1]{\sloppy\textcolor{red}{\sout{#1}}}    % delete

@@

However, ReSpecTh experimental files are another
XML-based format, and, as such, suffer from the same readability issues as PrIMe.
Moreover, \deletetwo{the standard remains inaccessible to the public, requiring potential users
to register with the standard's authors using a valid institutional email address
to access the specification and the growing database}
\addtwo{potential users must register using a valid institutional email address
to access the growing database}.

Nonetheless, databases of fundamental combustion experiments
have proved quite useful. For instance, Olm and colleagues used their
ReSpecTh-based database to quantify the performance of literature
hydrogen~\autocite{Olm:2014gn} and syngas~\autocite{Olm:2015ch} kinetic models,
and more recently develop improved models for \addtwo{hydrogen~\autocite{Varga:2015hy}},
hydrogen\slash syngas~\autocite{Varga:2016gj},
methanol\slash formaldehyde~\autocite{Olm:2017a},
and ethanol~\autocite{Olm:2016et}. In all \deletetwo{five}~\addtwo{six} cases, they converted numerous
experimental datasets from the literature into the ReSpecTh standard.
Their database currently contains \deletetwo{\num{93240}}~\addtwo{\num{93326}}
experimental data points in \deletetwo{1376}~\addtwo{1396}
ReSpecTh XML files~\autocite{respecth}.

@@

The second section of the file encodes the experimental data as a sequence of mappings. The
top-level key for the sequence is called \yabox|datapoints|. For the current version of the \ck{}
schema, only autoignition delay experiments are supported. The following information
\deleteone{os}~\addone{is} required in each element of the sequence:
%
\begin{itemize}
    \item \yabox|temperature| (required, sequence): The temperature of the
    experiment, with units and optionally uncertainty
    \item \yabox|ignition-delay| (required, sequence): The ignition delay of the
    experiment, with units and optionally uncertainty
    \item \yabox|pressure| (required, sequence): The pressure of the experiment,
    with units and optionally uncertainty
    \item \yabox|first-stage-ignition-delay| (optional, sequence): If two stages of ignition are
    present in the experiment, this is the length of the first stage of ignition, and the
    \yabox|ignition-delay| is then the overall ignition delay
    \item \yabox|equivalence-ratio| (optional, float): The value of the equivalence ratio
    \item \yabox|composition| (required, mapping): The composition of the
    mixture in the experiment, described via a nested mapping with the fields:
    \begin{itemize}
        \item \yabox|kind| (required, string): one of \yabox|mole fraction|, \yabox|mole percent|, or \yabox|mass fraction|
        \item \yabox|species|: sequence of mappings, with the fields:
        \begin{itemize}
            \item \yabox|species-name| (required, string): The name of the species
            \item \yabox|amount| (required, sequence): The mole fraction or percent, or the mass fraction, optionally with uncertainty
            \item \yabox|InChI|, \yabox|SMILES|, or \yabox|atomic-composition|
            (required, string or sequence): The InChI
            or SMILES string representing the molecule, or its atomic
            composition as a sequence
        \end{itemize}
    \end{itemize}
    \item \yabox|ignition-type| (required, mapping): The method used to detect
    ignition in the experiments. The required fields are
    \begin{itemize}
        \item \yabox|type| (required, string): How ignition delay was
        measured; one of \yabox|d/dt max| (indicates the ignition point was
        found at the maximum of the time derivative of the \yabox|target|),
        \yabox|max| or \yabox|min| (indicates the ignition point was at the
        maximum or minimum of the \yabox|target|), \yabox|1/2 max| (
        indicates the half-maximum point of \yabox|target|), or \yabox|d/dt max extrapolated|
        (indicates the maximum slope of the target extrapolated to the baseline)
        \item \yabox|target| (required, string): The target for the ignition
        \yabox|type| measurement; one of \yabox|temperature|, \yabox|pressure|,
        \yabox|OH|, \yabox|OH*|, \yabox|CH|, or \yabox|CH*|
    \end{itemize}
\end{itemize}

@@

Rapid compression machine experiments typically report several additional pieces of information in
addition to the ignition delay. As for shock tube experiments, these describe some of the
facility-specific effects during an experiment. The keys to describe this information are:
%
\begin{itemize}
    \item \yabox|volume-history| (optional, mapping): Specify the volume history of the reaction
    chamber as a function of time during a rapid compression machine experiment. The fields in this
    mapping are:
    \begin{itemize}
        \item \yabox|volume| (required, mapping): A mapping describing the volume in the history.
        The fields in this mapping are:
        \begin{itemize}
            \item \yabox|units| (required, string): the units of the volume, with dimensions length
            cubed
            \item \yabox|column| (required, integer): the \deletetwo{0-based} \addtwo{zero-based} index of the column storing the
            volume information in the \yabox|values| array \addtwo{(e.g., 0 or 1)}
        \end{itemize}
        \item \yabox|time| (required, mapping): A mapping describing the time in the history.
        The fields in this mapping are:
        \begin{itemize}
            \item \yabox|units| (required, string): the units of the time
            \item \yabox|column| (required, integer): the \deletetwo{0-based} \addtwo{zero-based} index of the column storing the
            time information in the \yabox|values| array \addtwo{(e.g., 0 or 1)}
        \end{itemize}
        \item \yabox|values| (required, sequence): A \deletetwo{sequence of sequences describing the values of
        the volume at the time points.}\addtwo{sequence of time-volume pairs describing the values
        of the volume at different times}
    \end{itemize}
    \item \yabox|compressed-temperature| (optional, sequence): The estimated temperature at the end
    of the compression stroke
    \item \yabox|compression-time| (optional, sequence): The time taken during the compression
    stroke
    \item \yabox|compressed-pressure| (optional, sequence): The measured pressure at the end of the
    compression stroke
\end{itemize}

@@

The \pybox|write_file()| method writes a new \ck{} YAML file based on the instance attributes
of the class, while \pybox|convert_to_ReSpecTh()| converts the \pybox|ChemKED|
instance to a ReSpecTh XML file. In the latter case, some information may be lost or the conversion
may not be possible, as \ck{} files support some data that ReSpecTh files do not support. We note
that the reverse is also true: not all ReSpecTh files may be
\addtwo{equivalently} converted to \ck{} files because not
all of the data that can be stored in a ReSpecTh format is (currently) supported by the \ck{} format.

>>>>

% For the version with changes, use these four commands
% *** Be very precise and careful about including whitespace and punctuation in your edits ***
% \usepackage[normalem]{ulem}
% \newcommand{\addone}[1]{{\sloppy\textcolor{teal}{\uline{#1}}}}  % add
% \newcommand{\deleteone}[1]{\sloppy\textcolor{teal}{\sout{#1}}}    % delete
% \newcommand{\addtwo}[1]{{\sloppy\textcolor{red}{\uline{#1}}}}  % add
% \newcommand{\deletetwo}[1]{\sloppy\textcolor{red}{\sout{#1}}}    % delete

@@

However, ReSpecTh experimental files are another
XML-based format, and, as such, suffer from the same readability issues as PrIMe.
Moreover, potential users must register using a valid institutional email address
to access the growing database.

Nonetheless, databases of fundamental combustion experiments
have proved quite useful. For instance, Olm and colleagues used their
ReSpecTh-based database to quantify the performance of literature
hydrogen~\autocite{Olm:2014gn} and syngas~\autocite{Olm:2015ch} kinetic models,
and more recently develop improved models for hydrogen~\autocite{Varga:2015hy},
hydrogen\slash syngas~\autocite{Varga:2016gj},
methanol\slash formaldehyde~\autocite{Olm:2017a},
and ethanol~\autocite{Olm:2016et}. In all six cases, they converted numerous
experimental datasets from the literature into the ReSpecTh standard.
Their database currently contains \num{93326}
experimental data points in 1396 ReSpecTh XML files~\autocite{respecth}.

@@

The second section of the file encodes the experimental data as a sequence of mappings. The
top-level key for the sequence is called \yabox|datapoints|. For the current version of the \ck{}
schema, only autoignition delay experiments are supported. The following information
is required in each element of the sequence:
%
\begin{itemize}
    \item \yabox|temperature| (required, sequence): The temperature of the
    experiment, with units and optionally uncertainty
    \item \yabox|ignition-delay| (required, sequence): The ignition delay of the
    experiment, with units and optionally uncertainty
    \item \yabox|pressure| (required, sequence): The pressure of the experiment,
    with units and optionally uncertainty
    \item \yabox|first-stage-ignition-delay| (optional, sequence): If two stages of ignition are
    present in the experiment, this is the length of the first stage of ignition, and the
    \yabox|ignition-delay| is then the overall ignition delay
    \item \yabox|equivalence-ratio| (optional, float): The value of the equivalence ratio
    \item \yabox|composition| (required, mapping): The composition of the
    mixture in the experiment, described via a nested mapping with the fields:
    \begin{itemize}
        \item \yabox|kind| (required, string): one of \yabox|mole fraction|, \yabox|mole percent|, or \yabox|mass fraction|
        \item \yabox|species|: sequence of mappings, with the fields:
        \begin{itemize}
            \item \yabox|species-name| (required, string): The name of the species
            \item \yabox|amount| (required, sequence): The mole fraction or percent, or the mass fraction, optionally with uncertainty
            \item \yabox|InChI|, \yabox|SMILES|, or \yabox|atomic-composition|
            (required, string or sequence): The InChI
            or SMILES string representing the molecule, or its atomic
            composition as a sequence
        \end{itemize}
    \end{itemize}
    \item \yabox|ignition-type| (required, mapping): The method used to detect
    ignition in the experiments. The required fields are
    \begin{itemize}
        \item \yabox|type| (required, string): How ignition delay was
        measured; one of \yabox|d/dt max| (indicates the ignition point was
        found at the maximum of the time derivative of the \yabox|target|),
        \yabox|max| or \yabox|min| (indicates the ignition point was at the
        maximum or minimum of the \yabox|target|), \yabox|1/2 max| (
        indicates the half-maximum point of \yabox|target|), or \yabox|d/dt max extrapolated|
        (indicates the maximum slope of the target extrapolated to the baseline)
        \item \yabox|target| (required, string): The target for the ignition
        \yabox|type| measurement; one of \yabox|temperature|, \yabox|pressure|,
        \yabox|OH|, \yabox|OH*|, \yabox|CH|, or \yabox|CH*|
    \end{itemize}
\end{itemize}

@@

Rapid compression machine experiments typically report several additional pieces of information in
addition to the ignition delay. As for shock tube experiments, these describe some of the
facility-specific effects during an experiment. The keys to describe this information are:
%
\begin{itemize}
    \item \yabox|volume-history| (optional, mapping): Specify the volume history of the reaction
    chamber as a function of time during a rapid compression machine experiment. The fields in this
    mapping are:
    \begin{itemize}
        \item \yabox|volume| (required, mapping): A mapping describing the volume in the history\footnote{Future versions
        of \ck{} will support specifying additional time histories for rapid compression machine and shock tube
        experiments (e.g., pressure, volume, light emission, OH emission). These will follow the specification format given for volume.}
        The fields in this mapping are:
        \begin{itemize}
            \item \yabox|units| (required, string): the units of the volume, with dimensions length
            cubed
            \item \yabox|column| (required, integer): the zero-based index of the column storing the
            volume information in the \yabox|values| array (e.g., 0 or 1)
        \end{itemize}
        \item \yabox|time| (required, mapping): A mapping describing the time in the history.
        The fields in this mapping are:
        \begin{itemize}
            \item \yabox|units| (required, string): the units of the time
            \item \yabox|column| (required, integer): the zero-based index of the column storing the
            time information in the \yabox|values| array (e.g., 0 or 1)
        \end{itemize}
        \item \yabox|values| (required, sequence): A sequence of time-volume pairs describing the values
        of the volume at different times
    \end{itemize}
    \item \yabox|compressed-temperature| (optional, sequence): The estimated temperature at the end
    of the compression stroke
    \item \yabox|compression-time| (optional, sequence): The time taken during the compression
    stroke
    \item \yabox|compressed-pressure| (optional, sequence): The measured pressure at the end of the
    compression stroke
\end{itemize}

@@

The \pybox|write_file()| method writes a new \ck{} YAML file based on the instance attributes
of the class, while \pybox|convert_to_ReSpecTh()| converts the \pybox|ChemKED|
instance to a ReSpecTh XML file. In the latter case, some information may be lost or the conversion
may not be possible, as \ck{} files support some data that ReSpecTh files do not support. We note
that the reverse is also true: not all ReSpecTh files may be
equivalently converted to \ck{} files because not
all of the data that can be stored in a ReSpecTh format is (currently) supported by the \ck{} format.
